Efficiency

### 2018-01-07 DraftExec.Rnw
for rosses eyes

The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.  In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.  Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate. However, in the case of a \$50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.  To that end, this paper has a very specific ambition.  It doesn't seek to advance the technical field with any new research or insight.  Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.

As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.

It should be noted that the technical fields involved here are vast and rapidly evolving.  The author makes no claim to be a world authority on the entirety of the topics which are touched. It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.

However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).  To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

@@

Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.  This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag, see \cite{carlton2007}  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the \textbf{distance} over which it is applied, not the time for which it is applied.  To think about it simply, gravity exerts a force on you downwards all the time.  But you don't expend any energy overcoming it when sitting still.  If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.

@@

A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.  Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element. At some local point for a moment in time, there is enough energy for the water present to boil.  It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam ciculating in the water.  It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.

>>>>

It is worthwhile explaining briefly the origins and intent of this paper, how it came to be written, and how it ought to be read.

The purpose is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

Having done my higher education in physics, and spent considerable time considering military systems running a small start-up defence company (Rubber Ducky Defence) I found the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs were obvious: propulsive efficiency.

It's definitely true that efficiency isn't everything in military technology.  For Rubber Ducky, I spent considerable time and money designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale (or even direction \parencite{davies2017}) of the impact to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance might be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.

The full translation this initial intrigue into the full substance of this paper has come about due to the determination and interest of Gary Johnson.  Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.

I should hasten to add that my expertise in this area is derived largely from an academic background which a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.  I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.  This paper doesn't seek to advance the technical field with any new research or insight.  It aims only to bring together as neatly and simply as possible those concepts and results which are neccessary and sufficient to give the appropriate weighting to this question.

As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found, and more thorough discussions of those topics which are relevant, but not essential to the question at hand.

It should be noted that the technical fields involved here are vast and rapidly evolving.  It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.

However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise embodiments in present design).  To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

@@

Perhaps the most important relationship to understand to comprehend the operation of submarines is the relationship between speed and drag .  This is important because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction to movement.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the \textbf{distance} over which it is applied (not the time for which it is applied).  To think about it simply, gravity exerts a force on you downwards all the time.  But you don't expend any energy overcoming it when sitting still.  If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.

@@

A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.  Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element. At some local point for a moment in time, there is enough energy for the water present to boil.  It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam circulating in the water.  It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.
